---
title: "Computing environment and similar model fits in R"
format: pdf
author:
  - name: Douglas Bates
    email: dmbates@gmail.com
    orcid: 0000-0001-8316-9503
    affiliation:
      - name: University of Wisconsin--Madison
        city: University of Wisconsin--Madison
        state: WI
        url: https://www.wisc.edu
        department: Department of Statistics
  - name: Phillip M. Alday
    email: phillip@phillipalday.com
    orcid: 0000-0002-9984-5745
    affiliation:
      - name: Beacon Biosignals
        city: Beacon Biosignals
        url: https://beacon.bio
  - name: Ajinkya H. Kokandakar
    email: ajinkya@stat.wisc.edu
    orcid: 0000-0001-6628-2272
    affiliations:
      - name: University of Wisconsin--Madison
        city: University of Wisconsin--Madison
        state: WI
        url: https://www.wisc.edu
        department: Department of Statistics
monofont: JuliaMono
date: last-modified
number-sections: true
execute:
  cache: true
engine: julia
julia:
  exeflags:
    - -tauto
    - --project=@.
---

# Computing environment

The timings in the paper, with the exception of the fits of models to the `ml-32m` data, were performed on a laptop with the following characteristics

```{julia}
using AppleAccelerate, InteractiveUtils, LinearAlgebra, RCall
versioninfo()
```

The versions of the Julia packages in use were

```{julia}
import Pkg
Pkg.status()
```

and the accelerated BLAS used for this computer were

```{julia}
BLAS.get_config()
```

# Timing similar model fits with lme4

For comparison we fit the same model as in section 3.1.1 using the `lme4::lmer` and `glmmTMB::tmbmod` functions in `R`

```{r}
dat <- lme4::InstEval
dat$service <- as.numeric(dat$service) - 1  # convert to a 0/1 numeric vector
ctrl <- lme4::lmerControl(calc.derivs = FALSE)
form <- y ~ 1 + service + (1 | s) + (1 | d) + (1 | dept) + (0 + service | dept)
mlmer <- lme4::lmer(form, dat, REML = FALSE, control = ctrl)
print(summary(mlmer, correlation = FALSE))
NULL       # suppress printing of the internal structure of the summary result
```

```{r}
tmbmod <- glmmTMB::glmmTMB(form, dat)
print(summary(tmbmod))
NULL
```

These model fits are essentially the same as that from `MixedModels.jl` but they take longer

```{r}
#| warning: false
print(
  microbenchmark::microbenchmark(
    lmer = lme4::lmer(form, dat, REML = FALSE, control = ctrl),
    glmmTMB = glmmTMB::glmmTMB(form, dat),
    times=6L,
    unit="seconds"
  )
)
NULL
```

Recall from section 3.1.1 of the paper that the `MixedModels.jl` fit of this model took just over one second on the same laptop computer.

We attempted to fit the `ml-32m` example using `glmmTMB` on a desktop computer with 64 GiB of memory but shut it down after it had been running for more than a week on the machine without converging.
Given otherwise similar performance, we do not expect `lme4` to perform any better than `glmmTMB`.

# Versions of R and the R packages that were used

```{r}
print(R.version.string)
print(packageVersion("lme4"))
print(packageVersion("glmmTMB"))
NULL
```

# Environment for the ml-32m model fits

The environment for the ml-32m model fits is shown in the `config.txt` file.  The Julia source code to download the `ml-32m.zip` file, extract the `ratings` table, create `moviesnratings.arrow` and `usersnratings.arrow`, followed by `ratings.arrow` is `ml32-download.jl`.

The Julia source file to run the `ml-32m` fits is `ml32-timefit.jl`.
Before you start these fits remember that they take up to a day on a powerful server computer.
Don't do this frivolously.
